---
title: "week7"
author: "Nicolas Legrand"
date: "2024-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
library(tidyverse)
set.seed(1000)

alpha <- 18
beta <- 2
n_obs <- 500

x <- runif(n_obs, 0, 30)

y <- alpha + beta*x + rnorm(n_obs, 0, 3)

data <- tibble(x = x, y = y)

ggplot(data, aes(x = x, y = y)) + geom_point(size = 2) + theme_bw() + labs(title = 'Simulated Data')
```


```{r}
ols_fit <- lm(y ~ x, data = data)
```

```{r}
summary(ols_fit)
```

```{r}
ols_fit$coefficients
```
TODO: create a cost function that returns the sum of squared errors.

```{r}
cost_function <- function(alpha, beta, x, y){
  ss <- 
  return(ss)
}
```


TODO: Compute the SSE of the model by inspecting the ols_fit variable looking at the residuals.

```{r}

```

Compute the MSE of the model by using the cost function we have created above.


```{r}
cost_function(
  alpha = ols_fit$coefficients[1][[1]],
  beta = ols_fit$coefficients[2][[1]],
  x = data$x, 
  y = data$y)
```

Plot the cost function across different values of beta.

```{r}
simulations = c()

for (i in -30:30){
  mse = cost_function(
  alpha = 18,
  beta = i,
  x = data$x, 
  y = data$y
  )
  
  simulations = c(simulations, mse)
}

```


```{r}
plot(x=-30:30, y=simulations)
```
Create a function gradient_desc that compute the gradient of the model depending on beta.

```{r}
gradient_desc <- function(alpha, beta, x, y){
  grad_beta <- 
  return (grad_beta)
}
```


```{r}
gradient_desc(
  alpha = 18,
  beta = 0.0,
  x = data$x, 
  y = data$y
)
```
TODO: create a gradient descent step.

```{r}
delta <- 0.001

minimize_function <- function(alpha, beta, x, y, delta){
  new_beta <- 
  return(new_beta)
}
```


```{r}
iter <- 100
res <- c(0)

for (i in 2:iter){
  this_beta = res[i-1]
  res[i] <- minimize_function(
    alpha = alpha, beta = this_beta, x = data$x, y = data$y, delta = delta
  )
}

res <- lapply(res, function(x) as.data.frame(t(x))) %>% bind_rows()
colnames(res) <- c('beta')

loss <- res %>% as_tibble() %>% rowwise() %>%
  summarise(sse = cost_function(18, beta, data$x, data$y))

res <- res %>% bind_cols(loss) %>%
  mutate(iteration = seq(1, 100)) %>% as_tibble()

res %>% glimpse()
```

```{r}
ggplot(res, aes(x = iteration, y = sse)) + geom_point(size = 2) + 
  theme_classic() + geom_line(aes(group = 1)) 
```
```{r}
ggplot(data, aes(x = x, y = y)) + 
  geom_point(size = 2) + 
  geom_abline(aes(intercept = alpha, slope = beta),
              data = res, size = 0.5, color = 'red') + 
  theme_classic() + 
  geom_abline(aes(intercept = alpha, slope = beta), 
              data = res %>% slice_head(), size = 0.5, color = 'blue') + 
  geom_abline(aes(intercept = alpha, slope = beta), 
              data = res %>% slice_tail(), size = 0.5, color = 'green')
```

