---
title: "Untitled"
output: html_document
date: "2024-02-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Example of a function
pernille_and_nicolas_function <- function(number){
  print("heyyy im calculating stuff wait a moment")
  new_number = number + 2
  return(new_number)
}

# See what happens
pernille_and_nicolas_function(2)

```

### Using R, plot the functions ð‘“1 and ð‘“2 in the range -5.0 to 5.0 as described above using the following parameters: ðœ‡ = 0.0, ðœŽ = 1.0, ðœ = 1.0. Can you spot any difference?

```{r}
# # Parameters
# mu <- 0.0
# sigma <-  1.0
# tau <-  1.0

# Define f1
f1 <- function(x, mu, sigma) {
  (1 / (sigma * sqrt(2 * pi))) * exp(-0.5 * ((x - mu) / sigma) ^ 2)
}

# Define f2
f2 <- function(x, mu, tau) {
  (sqrt(tau / (2 * pi))) * exp((-tau * (x - mu) ^ 2) / 2)
}

# Generate some x-values that we can put into f1 and f2
x_values <- seq(-5,
                5,
                by = 0.1)

# Let's calculate some y values using the functions f1 and f2 on our generated x's
y_f1 <- f1(x = x_values, 
           mu = 0.0,
           sigma = 3.0)

y_f2 <- f2(x = x_values,
           tau = 3.0,
           mu = 0
           )

# Create a dataframe for plotting
df <- data.frame(x = x_values,
                 f1 = y_f1,
                 f2 = y_f2)
# Let's plot
library(ggplot2)
plot1 <- ggplot(df,
                aes(x))+
  geom_line(aes(y = f1, colour = "f1"))+
  geom_line(aes(y = f2, colour = "f2"))+
  labs(title = "Probability density functions with mu = 0, sigma = 1, tau = 1",
       subtitle = "Learning about f1 and f2",
       caption = "Methods 2 #fun",
       colour = "Function")+
  theme_minimal()

plot1

# Let's create a col to check if theyre the same
library(tidyverse)
df <- df %>% mutate(checking_stuff = ifelse(f1 == f2, TRUE, FALSE))


# shift + ctrl + a on windows
# shift + cmd + a on mac (to clean up)


```
### Do the same, but this time using the following parameters: ðœ‡ = 0.0, ðœŽ = 3.0, ðœ = 3.0. How would you describe the influence of ðœ and ðœŽ on the width of the distribution?
```{r}
# It's done above
```

Higher tau --> higher precision --> narrower curve. Lower tau opposite.
Higher sigma --> higher standard deviatin --> wider curve. Lower sigma oposite.














